\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-10-26 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
newest changes sb

In this section, we illustrate the main functions of the \pkg{casebase} package through three case studies. Each one showcases a different type of analysis. First, we show how to model time flexibly as well as non-proportional hazards. Then we perform a competing-risk analysis and compare our results with the Cox model and the Fine-Gray model. The third case study illustrates how to perform variable selection in high-dimensional datasets.

## Case study 1---Modeling different functions of time and non-proportional hazards

@@

Note that we did not have to specify any cut points, as would be the case with the \code{survSplit} function in the \pkg{survival} package. 
The output object from \code{fitSmoothHazard} inherits from the \code{singleEventCB} and \code{glm} classes.
For this reason, we can leverage the \code{summary} method for \code{glm} objects to output a familiar summary of the results:

@@

Similarly, to compare different models (e.g. time modeled linearly), we could compute Akaike's Information Criterion (AIC) for each model and compare them.

@@

The \code{plot} method with \code{type="hr"} for objects of class \code{singleEventCB} can be used to compute time-dependent hazard ratios and confidence intervals. 
In Figure \ref{fig:interaction-ERSPC}, we show the estimated hazard ratio and 95% confidence interval for screening vs. control group as a function of time. 
Note that we must specify the covariate profiles and times for the predicted hazards.

@@

```{r interaction-ERSPC, eval = eval_cs1, fig.width=8, fig.height=6, fig.cap='Estimated hazard ratio and 95\\% confidence interval for screening vs. control group as a function of time in the ERSPC dataset. Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic pspline basis of follow-up time and treatment arm. 95\\% confidence intervals are calculated using the delta method. The plot shows that the effect of screening only begins to become statistically apparent by year 7. The 25-60\\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\\% reported in the original article.'}
new_time <- seq(1, 12, by  = 0.1)
new_data <- data.frame(ScrArm = factor("Control group",
                                         levels = c("Control group","Screening group")),
                      Follow.Up.Time = new_time)

@@

```{r cs1hazard, eval=eval_cs1, echo=TRUE}
plot(fit, type = "hazard",
     hazard.params = list(xvar = "Follow.Up.Time",
                          by = "ScrArm"))
```

@@

```{r, eval=FALSE, echo=FALSE}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- c(0, 1)

absoluteRisk(fit, time = new_time, newdata = new_data)
```

```{r, eval=eval_cs1, echo=TRUE}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0,14,0.1)
risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
head(risk)
```

@@

```{r erspc-cox-cif, eval = FALSE, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk functions for control and screening groups in the ERSPC data. We plot the estimate from the Cox model using \\code{survival::survfit} (solid line) and the estimate from case-base sampling (dashed line). We used a cubic B-spline expansion of time."}

new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0,14,0.1)
# Recompute with more time points
risk_mp <- absoluteRisk(fit, time = new_time, newdata = new_data)

plot(survfit(cox_model, newdata = new_data),
     ylab = "Risk probability (%)",
     xlab = "Years since randomization",
     fun = "event",
     xmax = max(new_time),
     conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100,
     xaxt = 'n')
legend("topleft", 
       legend = c("Control (Cox)","Control (Case-base)",
                  "Screening (Cox)", "Screening (Case-base)"), 
       col = rep(myCols, each = 2),
       lty = c(1, 2, 1, 2), 
       lwd = 2,
       cex = 1.1,
       bg = "gray90")
lines(risk_mp[,1], risk_mp[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risk_mp[,1], risk_mp[,3], col = myCols[2], lty = 2, lwd = 2)
```

@@

We would like to thank the anonymous reviewers for their insightful comments and criticisms. We would also like to thank Yi Yang for helpful discussions on penalized regression models. \mbox{Bhatnagar} and \mbox{Turgeon} both gratefully acknowledge funding via a Discovery Grant from the Natural Sciences and Engineering Research Council of Canada (NSERC).

>>>>

In this section, we illustrate the main functions of the \pkg{casebase} package through three case studies. Each one showcases a different type of analysis. First, we show how to model non-constant and non-proportional hazards through a flexible specification of time. Then we perform a competing-risk analysis and compare our results with the Cox model and the Fine-Gray model. The third case study illustrates how to perform variable selection in high-dimensional datasets.

## Case study 1---Flexible modeling of the hazard function

@@

<!--Note that we did not have to specify any cut points, as would be the case 
with the \code{survSplit} function in the \pkg{survival} package.--> 
The output object from \code{fitSmoothHazard} inherits from the \code{singleEventCB} 
and \code{glm} classes. For this reason, we can leverage the \code{summary} 
method for \code{glm} objects to output a familiar summary of the results:

@@

Similarly, to compare different models (e.g. time modeled linearly), we could 
compute Akaike's Information Criterion (AIC) for each model.

@@

The \code{plot} method with \code{type="hr"} for objects of class 
\code{singleEventCB} can be used to compute time-dependent hazard ratios and 
confidence intervals. In Figure \ref{fig:interaction-ERSPC}, we show the 
estimated hazard ratio and 95% confidence interval for screening vs. control 
group as a function of time. Note that we must specify the covariate profile for 
the reference group and times for the predicted hazards.

@@

```{r interaction-ERSPC, eval = eval_cs1, fig.width=8, fig.height=6, fig.cap='Estimated hazard ratio and 95\\% confidence interval for screening vs. control group as a function of time in the ERSPC dataset. Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic pspline basis (df=2) of follow-up time and treatment arm. 95\\% confidence intervals are calculated using the delta method. The plot shows that the effect of screening only begins to become statistically apparent by year 7. The 25-60\\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\\% reported in the original article.'}
new_time <- seq(1, 12, by  = 0.1)
new_data <- data.frame(ScrArm = factor("Control group",
                                         levels = c("Control group","Screening group")),
                      Follow.Up.Time = new_time)

@@

```{r cs1hazard, eval=eval_cs1, echo=TRUE, fig.cap='Estimated hazard functions for control and screening groups in the ERSPC dataset. Hazards are estimated from fitting a parametric model with casebase sampling as a function of the interaction between a cubic pspline basis (df=2) of follow-up time and treatment arm. The package vignettes provide a detailed description of how to plot hazard functions for any combination of covariates along with confidence bands.'}
plot(fit, type = "hazard",
     hazard.params = list(xvar = "Follow.Up.Time",
                          by = "ScrArm"))
```

\FloatBarrier

@@

```{r, eval=eval_cs1, echo=TRUE}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0,14,0.1)
risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
```

@@

@@

We would like to thank the anonymous reviewers for their insightful comments and criticisms. We would also like to thank Yi Yang for helpful discussions on penalized regression models. \mbox{Bhatnagar} (RGPIN-2020-05133) and \mbox{Turgeon} (RGPIN-2021-04073) both gratefully acknowledge funding via a Discovery Grant from the Natural Sciences and Engineering Research Council of Canada (NSERC), RGPIN.

### 2021-10-26 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
newest changes sb

In this section, we illustrate the main functions of the \pkg{casebase}
package through three case studies. Each one showcases a different type
of analysis. First, we show how to model time flexibly as well as
non-proportional hazards. Then we perform a competing-risk analysis and
compare our results with the Cox model and the Fine-Gray model. The
third case study illustrates how to perform variable selection in
high-dimensional datasets.

\hypertarget{case-study-1modeling-different-functions-of-time-and-non-proportional-hazards}{%
\subsection{Case study 1---Modeling different functions of time and
non-proportional
hazards}\label{case-study-1modeling-different-functions-of-time-and-non-proportional-hazards}}

@@

Note that we did not have to specify any cut points, as would be the
case with the \code{survSplit} function in the \pkg{survival} package.
The output object from \code{fitSmoothHazard} inherits from the
\code{singleEventCB} and \code{glm} classes. For this reason, we can
leverage the \code{summary} method for \code{glm} objects to output a
familiar summary of the results:

@@

Similarly, to compare different models (e.g.~time modeled linearly), we
could compute Akaike's Information Criterion (AIC) for each model and
compare them.

@@

The \code{plot} method with \code{type="hr"} for objects of class
\code{singleEventCB} can be used to compute time-dependent hazard ratios
and confidence intervals. In Figure \ref{fig:interaction-ERSPC}, we show
the estimated hazard ratio and 95\% confidence interval for screening
vs.~control group as a function of time. Note that we must specify the
covariate profiles and times for the predicted hazards.

@@

@@

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./interaction-ERSPC-1} \caption[Estimated hazard ratio and 95\% confidence interval for screening vs]{Estimated hazard ratio and 95\% confidence interval for screening vs. control group as a function of time in the ERSPC dataset. Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic pspline basis of follow-up time and treatment arm. 95\% confidence intervals are calculated using the delta method. The plot shows that the effect of screening only begins to become statistically apparent by year 7. The 25-60\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\% reported in the original article.}\label{fig:interaction-ERSPC}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
plot(fit, type = "hazard",
     hazard.params = list(xvar = "Follow.Up.Time",
                          by = "ScrArm"))
\end{Sinput}
\begin{Soutput}
#> Conditions used in construction of plot
#> ScrArm: Control group / Screening group
#> offset: 0
\end{Soutput}

\includegraphics[width=\textwidth,keepaspectratio=true]{./cs1hazard-1} \end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0,14,0.1)
risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
head(risk)
\end{Sinput}
\begin{Soutput}
#>  time                
#>   0.0 0.0e+00 0.0e+00
#>   0.1 2.0e-06 6.2e-06
#>   0.2 4.2e-06 1.2e-05
#>   0.3 6.6e-06 1.8e-05
#>   0.4 9.2e-06 2.3e-05
#>   0.5 1.2e-05 2.9e-05
\end{Soutput}
\end{Schunk}

>>>>

In this section, we illustrate the main functions of the \pkg{casebase}
package through three case studies. Each one showcases a different type
of analysis. First, we show how to model non-constant and
non-proportional hazards through a flexible specification of time. Then
we perform a competing-risk analysis and compare our results with the
Cox model and the Fine-Gray model. The third case study illustrates how
to perform variable selection in high-dimensional datasets.

\hypertarget{case-study-1flexible-modeling-of-the-hazard-function}{%
\subsection{Case study 1---Flexible modeling of the hazard
function}\label{case-study-1flexible-modeling-of-the-hazard-function}}

@@

The output object from \code{fitSmoothHazard} inherits from the
\code{singleEventCB} and \code{glm} classes. For this reason, we can
leverage the \code{summary} method for \code{glm} objects to output a
familiar summary of the results:

@@

Similarly, to compare different models (e.g.~time modeled linearly), we
could compute Akaike's Information Criterion (AIC) for each model.

@@

The \code{plot} method with \code{type="hr"} for objects of class
\code{singleEventCB} can be used to compute time-dependent hazard ratios
and confidence intervals. In Figure \ref{fig:interaction-ERSPC}, we show
the estimated hazard ratio and 95\% confidence interval for screening
vs.~control group as a function of time. Note that we must specify the
covariate profile for the reference group and times for the predicted
hazards.

@@

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./interaction-ERSPC-1} \caption[Estimated hazard ratio and 95\% confidence interval for screening vs]{Estimated hazard ratio and 95\% confidence interval for screening vs. control group as a function of time in the ERSPC dataset. Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic pspline basis (df=2) of follow-up time and treatment arm. 95\% confidence intervals are calculated using the delta method. The plot shows that the effect of screening only begins to become statistically apparent by year 7. The 25-60\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\% reported in the original article.}\label{fig:interaction-ERSPC}
\end{figure}
\end{Schunk}

@@

@@

\begin{Schunk}
\begin{Sinput}
plot(fit, type = "hazard",
     hazard.params = list(xvar = "Follow.Up.Time",
                          by = "ScrArm"))
\end{Sinput}
\begin{Soutput}
#> Conditions used in construction of plot
#> ScrArm: Control group / Screening group
#> offset: 0
\end{Soutput}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./cs1hazard-1} \caption[Estimated hazard functions for control and screening groups in the ERSPC dataset]{Estimated hazard functions for control and screening groups in the ERSPC dataset. Hazards are estimated from fitting a parametric model with casebase sampling as a function of the interaction between a cubic pspline basis (df=2) of follow-up time and treatment arm. The package vignettes provide a detailed description of how to plot hazard functions for any combination of covariates along with confidence bands.}\label{fig:cs1hazard}
\end{figure}
\end{Schunk}

\FloatBarrier

@@

\begin{Schunk}
\begin{Sinput}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0,14,0.1)
risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
\end{Sinput}
\end{Schunk}
